---
title: "Hands-on Exercise 4b: Visual Statistical Analysis"
date: February 1,2024
date-modified: "last-modified"
toc: true
execute: 
  eval: true
  echo: true
  warning: false
code-fold: true
code-summary: "Show the code"
---
# 1 Getting Started

In this exercise, we will use the following our R packages.

-   [tidyverse](https://www.tidyverse.org/), a family of R packages for data science process,

-   [ggstatsplot ](https://indrajeetpatil.github.io/ggstatsplot/), a ggplot2 extension created graphics with details from statistical tests included in the information-rich plots themselves.


::: panel-tabset
## Loading R package

The code chunk below uses `p_load()` of pacman package to check if these packages are installed in the computer and load them onto your working R environment.

```{r}
#| code-fold: false
pacman::p_load(ggstatsplot, tidyverse) 
```

## Importing Data

The code chunk below imports exam_data.csv into R environment by using `read_csv()` function of [readr](https://readr.tidyverse.org/) package.

```{r}
#| code-fold: false
exam <- read.csv("data/Exam_data.csv")
```

## Summarizing Data

The code chunk below uses `summary()`to summarize the data.

```{r}
#| code-fold: false
summary(exam)
```
:::