---
title: "Hands-on Exercise 4c: Visualizing Uncertainty"
date: February 1,2024
date-modified: "last-modified"
toc: true
execute: 
  eval: true
  echo: true
  warning: false
code-fold: true
code-summary: "Show the code"
---

# 1 Getting Started

In this exercise, we will use the following our R packages.

-   tidyverse, a family of R packages for data science process,

-   plotly for creating interactive plot,

-   gganimate for creating animation plot,

-   DT for displaying interactive html table,

-   crosstalk for for implementing cross-widget interactions (currently, linked brushing and filtering), and

-   ggdist for visualising distribution and uncertainty.

::: panel-tabset
## Loading R package

The code chunk below uses `p_load()` of pacman package to check if these packages are installed in the computer and load them onto your working R environment.

```{r}
#| code-fold: false
devtools::install_github("wilkelab/ungeviz")
```

```{r}
#| code-fold: false
pacman::p_load(ungeviz, plotly, crosstalk,
               DT, ggdist, ggridges,
               colorspace, gganimate, tidyverse)
```

## Importing Data

The code chunk below imports exam_data.csv into R environment by using `read_csv()` function of [readr](https://readr.tidyverse.org/) package.

```{r}
#| code-fold: false
exam <- read.csv("data/Exam_data.csv")
```

## Summarizing Data

The code chunk below uses `summary()`to summarize the data.

```{r}
#| code-fold: false
summary(exam)
```
:::

# 2 Visualizing the uncertainty of point estimates: ggplot2 methods

A point estimate is a single number, such as a mean. Uncertainty, on the other hand, is expressed as standard error, confidence interval, or credible interval.

::: callout-important
## Important

-   Don’t confuse the uncertainty of a point estimate with the variation in the sample
:::

In this section, you will learn how to plot error bars of maths scores by race by using data provided in *exam* tibble data frame.

Firstly, code chunk below will be used to derive the necessary summary statistics.

```{r}
#| code-fold: false
my_sum <- exam %>%
  group_by(RACE) %>%
  summarise(
    n=n(),
    mean=mean(MATHS),
    sd=sd(MATHS)
    ) %>%
  mutate(se=sd/sqrt(n-1))
```

::: callout-tip
## Things to learn from the code chunk above

-   `group_by()` of **dplyr** package is used to group the observation by RACE,

-   `summarise()` is used to compute the count of observations, mean, standard deviation

-   `mutate()` is used to derive standard error of Maths by RACE, and

-   the output is save as a tibble data table called *my_sum*.
:::

Next, the code chunk below will be used to display *my_sum* tibble data frame in an html table format.

::: panel-tabset
## Code chunk

```{r}
#| eval: false 
#| code-fold: false
knitr::kable(head(my_sum), format = 'html')
```

## Table

```{r}
#| echo: false
knitr::kable(head(my_sum), format = 'html')
```
:::

## 2.1 Plotting standard error bars of point estimates

The below plot shows the standard error bars of mean maths score by race.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(my_sum) +
  geom_errorbar(
    aes(x=RACE, 
        ymin=mean-se, 
        ymax=mean+se), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    size=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  ggtitle("Standard error of mean maths score by race")
```

## Code chunk

```{r}
#| eval: false 
#| code-fold: false
ggplot(my_sum) +
  geom_errorbar(
    aes(x=RACE, 
        ymin=mean-se, 
        ymax=mean+se), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    size=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  ggtitle("Standard error of mean maths score by race")
```

::: callout-tip
## Things to learn from the code chunk above

-   The error bars are computed by using the formula mean+/-se.

-   For `geom_point()`, it is important to indicate stat=“identity”.
:::
:::

## 2.2 Plotting confidence interval of point estimates

The below plot shows the confidence intervals of mean maths score by race.

::: panel-tabset
## Plot

```{r}
#| echo: false
ggplot(my_sum) +
  geom_errorbar(
    aes(x=reorder(RACE, -mean), 
        ymin=mean-1.96*se, 
        ymax=mean+1.96*se), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    size=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  labs(x = "Maths score",
       title = "95% confidence interval of mean maths score by race")
```

## Code chunk

```{r}
#| eval: false 
#| code-fold: false
ggplot(my_sum) +
  geom_errorbar(
    aes(x=reorder(RACE, -mean), 
        ymin=mean-1.96*se, 
        ymax=mean+1.96*se), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    size=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  labs(x = "Maths score",
       title = "95% confidence interval of mean maths score by race")
```

::: callout-tip
## Things to learn from the code chunk above

-   The confidence intervals are computed by using the formula mean+/-1.96\*se.

-   The error bars is sorted by using the average maths scores.

-   `labs()` argument of ggplot2 is used to change the x-axis label.
:::
:::

## 2.3 Visualizing the uncertainty of point estimates with interactive error bars

In this section, we will show how to plot interactive error bars for the 99% confidence interval of mean maths score by race as shown in the figure below.

::: panel-tabset
## Plot

```{r}
#| echo: false
shared_df = SharedData$new(my_sum)

bscols(widths = c(4,8),
       ggplotly((ggplot(shared_df) +
                   geom_errorbar(aes(
                     x=reorder(RACE, -mean),
                     ymin=mean-2.58*se, 
                     ymax=mean+2.58*se), 
                     width=0.2, 
                     colour="black", 
                     alpha=0.9, 
                     size=0.5) +
                   geom_point(aes(
                     x=RACE, 
                     y=mean, 
                     text = paste("Race:", `RACE`, 
                                  "<br>N:", `n`,
                                  "<br>Avg. Scores:", round(mean, digits = 2),
                                  "<br>95% CI:[", 
                                  round((mean-2.58*se), digits = 2), ",",
                                  round((mean+2.58*se), digits = 2),"]")),
                     stat="identity", 
                     color="red", 
                     size = 1.5, 
                     alpha=1) + 
                   xlab("Race") + 
                   ylab("Average Scores") + 
                   theme_minimal() + 
                   theme(axis.text.x = element_text(
                     angle = 45, vjust = 0.5, hjust=1)) +
                   ggtitle("99% Confidence interval of average /<br>maths scores by race")), 
                tooltip = "text"), 
       DT::datatable(shared_df, 
                     rownames = FALSE, 
                     class="compact", 
                     width="100%", 
                     options = list(pageLength = 10,
                                    scrollX=T), 
                     colnames = c("No. of pupils", 
                                  "Avg Scores",
                                  "Std Dev",
                                  "Std Error")) %>%
         formatRound(columns=c('mean', 'sd', 'se'),
                     digits=2))
```

## Code chunk

```{r}
#| eval: false 
#| code-fold: false
shared_df = SharedData$new(my_sum)

bscols(widths = c(4,8),
       ggplotly((ggplot(shared_df) +
                   geom_errorbar(aes(
                     x=reorder(RACE, -mean),
                     ymin=mean-2.58*se, 
                     ymax=mean+2.58*se), 
                     width=0.2, 
                     colour="black", 
                     alpha=0.9, 
                     size=0.5) +
                   geom_point(aes(
                     x=RACE, 
                     y=mean, 
                     text = paste("Race:", `RACE`, 
                                  "<br>N:", `n`,
                                  "<br>Avg. Scores:", round(mean, digits = 2),
                                  "<br>95% CI:[", 
                                  round((mean-2.58*se), digits = 2), ",",
                                  round((mean+2.58*se), digits = 2),"]")),
                     stat="identity", 
                     color="red", 
                     size = 1.5, 
                     alpha=1) + 
                   xlab("Race") + 
                   ylab("Average Scores") + 
                   theme_minimal() + 
                   theme(axis.text.x = element_text(
                     angle = 45, vjust = 0.5, hjust=1)) +
                   ggtitle("99% Confidence interval of average /<br>maths scores by race")), 
                tooltip = "text"), 
       DT::datatable(shared_df, 
                     rownames = FALSE, 
                     class="compact", 
                     width="100%", 
                     options = list(pageLength = 10,
                                    scrollX=T), 
                     colnames = c("No. of pupils", 
                                  "Avg Scores",
                                  "Std Dev",
                                  "Std Error")) %>%
         formatRound(columns=c('mean', 'sd', 'se'),
                     digits=2))
```
:::

# 3 Visualizing Uncertainty: ggdist package

-   [**ggdist**](https://mjskay.github.io/ggdist/) is an R package that provides a flexible set of ggplot2 geoms and stats designed especially for visualising distributions and uncertainty.

-   It is designed for both frequentist and Bayesian uncertainty visualization, taking the view that uncertainty visualization can be unified through the perspective of distribution visualization:

    -   for frequentist models, one visualises confidence distributions or bootstrap distributions (see vignette(“freq-uncertainty-vis”));

    -   for Bayesian models, one visualises probability distributions (see the tidybayes package, which builds on top of ggdist).

## 3.1 Visualizing the uncertainty of point estimates: ggdist methods

In the code chunk below, [`stat_pointinterval()`](https://mjskay.github.io/ggdist/reference/stat_pointinterval.html) of **ggdist** is used to build a visual for displaying distribution of maths scores by race.

```{r}
exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_pointinterval(.width = 0.95,
                     .point = median,
                     .interval = qi) +
  labs(title = "Visualizing confidence intervals of mean math score",
    subtitle = "Mean Point + Multiple-interval plot")
```

In the code chunk below, [`stat_gradientinterval()`](https://mjskay.github.io/ggdist/reference/stat_gradientinterval.html) of **ggdist** is used to build a visual for displaying distribution of maths scores by race.

```{r}
exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_gradientinterval(   
    fill = "skyblue",      
    show.legend = TRUE     
  ) +                        
  labs(
    title = "Visualizing confidence intervals of mean math score",
    subtitle = "Gradient + interval plot")
```

# 4  Visualizing Uncertainty with Hypothetical Outcome Plots (HOPs)

Launch the application in R
```{r}
#| code-fold: false
library(ungeviz)

```

```{r}
ggplot(data = exam, 
       (aes(x = factor(RACE), 
            y = MATHS))) +
  geom_point(position = position_jitter(
    height = 0.3, 
    width = 0.05), 
    size = 0.4, 
    color = "#0072B2", 
    alpha = 1/2) +
  geom_hpline(data = sampler(25, 
                             group = RACE), 
              height = 0.6, 
              color = "#D55E00") +
  theme_bw() + 
  transition_states(.draw, 1, 3)

```
