---
title: "Take-home Exercise 1: Creating data visualisation beyond default"
date: January 16,2024
date-modified: "last-modified"
toc: true
execute: 
  eval: true
  echo: true
  warning: false
  message: false
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

# 1 Overview

Andreas Schleicher, the director of OECD education,shared in a BBC article that “Singapore managed to achieve excellence without wide differences between children from wealthy and disadvantaged families.” (2016) Additionally, many Singapore's Minister for Education have adopted the slogan “every school a good school.” However, the perception among the public is that there are still disparities that exist, especially between the elite schools and neighborhood school, between students from families with higher socioeconomic status and those with relatively lower socioeconomic status and immigration and non-immigration families.

# 2 Objectives

The objective of this study is to use appropriate **Exploratory Data Analysis (EDA)** methods and **ggplot2** functions to reveal:

-   The distribution of Singapore students’ performance in mathematics, reading, and science, and

-   The relationship between these performances with schools, gender and socioeconomic status of the students.

# 3 Data Preparation

## 3.1 Installing R packages

The code chunk below uses `p_load()` of **pacman** package to check if multiple packages are installed in the computer. If they are, then they will be launched into R.

```{r}
#| code-fold: false
pacman::p_load(tidyverse, haven, ggrepel, patchwork,ggplot2, 
               ggthemes, hrbrthemes,dplyr,tidyr,knitr,ggridges)
```

## 3.2 Importing Data

The 2022 Programme for International Student Assessment (PISA) data released on December 5, 2022 will be used. PISA global education survey every three years to assess the education systems worldwide through testing 15 year old students in the subjects of mathematics, reading, and science.

The code chunk below uses [`read_sas()`](https://haven.tidyverse.org/reference/read_sas.html) of [haven](https://haven.tidyverse.org/index.html) to import PISA data into R environment.

```{r}
#| code-fold: false
stu_qqq <- read_sas("data/cy08msp_stu_qqq.sas7bdat")
```

::: panel-tabset
## Filtering data

Filter Singaporean data by using [`filter()`](https://dplyr.tidyverse.org/reference/filter.html#useful-filter-functions)

```{r}
#| code-fold: false
stu_qqq_SG <- stu_qqq %>%
  filter(CNT == "SGP")
```

## Selecting variables

The code chunk below uses [`select()`](https://dplyr.tidyverse.org/reference/select.html) to include selected variables.

```{r}
#| code-fold: false
stu_qqq_SG_v2 <- stu_qqq_SG %>%
  select(CNTSTUID,PV1MATH,PV1READ,PV1SCIE,STRATUM,ST004D01T,ESCS) %>%
  rename(STUDENT_ID=CNTSTUID,
         MATH=PV1MATH,
         READING=PV1READ,
         SCIENCE=PV1SCIE,
         SCHOOL=STRATUM,
         SOCIOECONOMIC_STATUS=ESCS,
         GENDER=ST004D01T)
```

::: callout-Note
According to [OECD](https://read.oecd-ilibrary.org/education/pisa-data-analysis-manual-sas-second-edition_9789264056251-en#page1), it stated that single plausible value can be used to analzye in the exploratory phase of the data. Using one or five plausible values will not make any significant difference. Therefore, we will proceed with PV1 only for further analysis.
:::
:::

## 3.3 Wrangling data

::: panel-tabset
## Data structure

The code chunk below uses `glimpse` to check the data structure

```{r}
#| code-fold: false
glimpse(stu_qqq_SG_v2)
```

## Duplicates

The code chunk below uses `duplicated()` to check duplicates

```{r}
#| code-fold: false
stu_qqq_SG_v2[duplicated(stu_qqq_SG_v2),]
```

From the result, there is no duplicate.

## Missing values

The code chunk below uses to check missing values

```{r}
#| code-fold: false
sum(is.na(stu_qqq_SG_v2))

```

The result shows that there are missing values

The code chunk below uses `drop_na()` to remove missing values

```{r}
#| code-fold: false
stu_qqq_SG_v3 <- stu_qqq_SG_v2 %>%
  drop_na()
sum(is.na(stu_qqq_SG_v3))
```

## Recoding and Binning

The code below use `recode()` to recode gender and `nitle()` to bin socioeconomic status

```{r}
#| code-fold: false
stu_qqq_SG_final <- stu_qqq_SG_v3 %>%
  mutate(GENDER = recode(GENDER,
                         `1` = "Female",
                         `2` = "Male"),
         SCHOOL = recode(SCHOOL,
                         "SGP01" = "Public School",
                         "SGP03" = "Private School")) %>%
  mutate(SOCIOECONOMIC_STATUS = ntile(SOCIOECONOMIC_STATUS, 4), .after = SOCIOECONOMIC_STATUS) %>%
  mutate(SOCIOECONOMIC_STATUS = recode(SOCIOECONOMIC_STATUS,
                                      `1` = "Highly Disadvantaged",
                                      `2` = "Disadvantaged",
                                      `3` = "Advantaged",
                                      `4` = "Highly Advantaged")) %>%
  mutate(SOCIOECONOMIC_STATUS = factor(SOCIOECONOMIC_STATUS, 
                                      levels = c("Highly Disadvantaged", "Disadvantaged", 
                                                 "Advantaged", "Highly Advantaged"),
                                      ordered = TRUE))
```
:::

# 4 Final Dataset

The final data is displayed by using `knitr::kable()` function

```{r}
#| code-fold: false
kable(head(stu_qqq_SG_final), "simple")
```

A statistical summary of final dataset by using `summary()`

```{r}
#| code-fold: false
summary(stu_qqq_SG_final)
```

# 5 Exploratory Data Analysis

## 5.1 Distribution of Students' Performance

The below panel is shown the distribution of Singapore students’ performance in mathematics, reading, and science.

::: panel-tabset
## MATHS

```{r}
#calculate Mean and Median 
mean_math <- mean(stu_qqq_SG_final$MATH,na.rm = TRUE)
median_math <-median(stu_qqq_SG_final$MATH,na.rm = TRUE)

ggplot(stu_qqq_SG_final, aes(x = MATH)) +
  geom_histogram(bins = 30, fill = "beige", color = "grey") +
  scale_x_continuous(limits = c(0, 1000)) + 
  labs(title ="Distribution of Math Scores", y="count") +
  geom_vline(xintercept = mean_math,
             color = "red", 
             linetype = "dashed", 
             size = 1) +
  annotate("text", x = 480, y = 900,
           label = paste("Mean =", round(mean_math, 2)), 
           color = "red", 
           size = 3.5)+  
  geom_vline(xintercept = median_math,
             color = "blue", 
             linetype = "dashed", 
             size = 1) +
  annotate("text", x = 700, y = 900,
           label = paste("Median =", round(median_math, 2)), 
           color = "blue", 
           size = 3.5)
```

## READING

```{r}
mean_reading <- mean(stu_qqq_SG_final$READING, na.rm = TRUE)
median_reading <- median(stu_qqq_SG_final$READING, na.rm = TRUE)

ggplot(stu_qqq_SG_final, aes(x = READING)) +
  geom_histogram(bins = 30, fill = "pink", color = "grey") +
  scale_x_continuous(limits = c(0, 1000)) + 
  labs(title ="Distribution of Reading Scores", y="count") +
  geom_vline(xintercept = mean_reading,
             color = "red", 
             linetype = "dashed", 
             size = 1) +
  annotate("text", x = 400, y = 900,
           label = paste("Mean =", round(mean_reading, 2)), 
           color = "red", 
           size = 3.5)+  
  geom_vline(xintercept = median_reading,
             color = "blue", 
             linetype = "dashed", 
             size = 1) +
  annotate("text", x = 700, y = 900,
           label = paste("Median =", round(median_reading, 2)), 
           color = "blue", 
           size = 3.5)
```

## SCIENCE

```{r}
mean_science <- mean(stu_qqq_SG_final$SCIENCE, na.rm = TRUE)
median_science <- median(stu_qqq_SG_final$SCIENCE, na.rm = TRUE)

ggplot(stu_qqq_SG_final, aes(x = SCIENCE)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "grey") +
  scale_x_continuous(limits = c(0, 1000)) +
  labs(title = "Distribution of Science Scores", y = "Count") +
  geom_vline(xintercept = mean_science,
             color = "red", 
             linetype = "dashed",
             size = 1) +
  annotate("text", x = 480, y = 900,  
           label = paste("Mean =", round(mean_science, 2)), 
           color = "red", 
           size = 3.5) +  
  geom_vline(xintercept = median_science,
             color = "blue",
             linetype = "dashed", 
             size = 1) +
  annotate("text", x = 700, y = 900,  
           label = paste("Median =", round(median_science, 2)), 
           color = "blue", 
           size = 3.5)
```
:::

::: callout-note
The distribution among maths, science, and reading appear to be quite similar.

However, when comparing all three subjects, maths has the highest values for both mean and median, while reading has the lowest mean and median scores.
:::

## 5.2 Performance and Gender

To find some insights between the performance of each subjects and gender, we will use density graph.

```{r}
p1 <- ggplot(data=stu_qqq_SG_final, 
       aes(x = MATH,
           color=GENDER,
           fill=GENDER)) +
  geom_density(linewidth = 0.05, alpha=0.6)+
  scale_fill_manual(values = c("#AB1858", "white")) +
  scale_color_manual(values = c("#AB1858", "grey"))+
  scale_x_continuous(limits = c(0, 1000))

p2 <- ggplot(data=stu_qqq_SG_final, 
       aes(x = READING,
           color=GENDER,
           fill=GENDER)) +
  geom_density(linewidth = 0.05, alpha=0.6)+
  scale_fill_manual(values = c("#AB1858", "white")) +
  scale_color_manual(values = c("#AB1858", "grey"))+
  scale_x_continuous(limits = c(0, 1000))

p3 <- ggplot(data=stu_qqq_SG_final, 
       aes(x = SCIENCE,
           color=GENDER,
           fill=GENDER)) +
  geom_density(linewidth = 0.05, alpha=0.6)+
  scale_fill_manual(values = c("#AB1858", "white")) +
  scale_color_manual(values = c("#AB1858", "grey"))+
  scale_x_continuous(limits = c(0, 1000))

(p1/p2/p3)


```

::: callout-note
The distributions of performance in each subject between gender are slightly different.

The density plot of females have a higher peak compared to males. This suggests that more females tend to have the similar scores, whereas males' score appear to be more scattered indicating a wider range of performance among them.
:::

## 5.3 Performance and School Type

In this section, we will use density graph and boxplots to find insights related to performance and type of schools

::: panel-tabset
## MATHS

```{r}

ggplot(data=stu_qqq_SG_final, 
       aes(y = MATH, x=SCHOOL,fill=SCHOOL)) +
  geom_boxplot()+
  scale_y_continuous(limits = c(0, 1000))+
  theme(legend.position = "none")+
  stat_summary(geom = "point",       
               fun.y="mean",         
               colour ="black",        
               size=2)

ggplot(data=stu_qqq_SG_final, 
       aes(x = MATH,
           color=SCHOOL,
           fill=SCHOOL)) +
  geom_density(linewidth = 0.05, alpha=0.6)+
  scale_fill_manual(values = c("#67B5F3", "white")) +
  scale_color_manual(values = c("#67B5F3", "grey"))+
  scale_x_continuous(limits = c(0, 1000))+
  theme(legend.position = c(0.85, 0.85))+
  labs(title = "Distribution of Math Scores across Different School Types")


```

## READING

```{r}
ggplot(data=stu_qqq_SG_final, 
       aes(y = READING, x=SCHOOL,fill=SCHOOL)) +
  geom_boxplot()+
  scale_y_continuous(limits = c(0, 1000))+
  theme(legend.position = "none")+
  stat_summary(geom = "point",       
               fun.y="mean",         
               colour ="black",        
               size=2)

ggplot(data=stu_qqq_SG_final, 
       aes(x = READING,
           color=SCHOOL,
           fill=SCHOOL)) +
  geom_density(linewidth = 0.05, alpha=0.6)+
  scale_fill_manual(values = c("#67B5F3", "white")) +
  scale_color_manual(values = c("#67B5F3", "grey"))+
  scale_x_continuous(limits = c(0, 1000))+
  theme(legend.position = c(0.85, 0.85))+
  labs(title = "Distribution of Reading Scores across Different School Types")

```

## SCIENCE

```{r}
ggplot(data=stu_qqq_SG_final, 
       aes(y = SCIENCE, x=SCHOOL,fill=SCHOOL)) +
  geom_boxplot()+
  scale_y_continuous(limits = c(0, 1000))+
  theme(legend.position = "none")+
  stat_summary(geom = "point",       
               fun.y="mean",         
               colour ="black",        
               size=2)

ggplot(data=stu_qqq_SG_final, 
       aes(x = SCIENCE,
           color=SCHOOL,
           fill=SCHOOL)) +
  geom_density(linewidth = 0.05, alpha=0.6)+
  scale_fill_manual(values = c("#67B5F3", "white")) +
  scale_color_manual(values = c("#67B5F3", "grey"))+
  scale_x_continuous(limits = c(0, 1000))+
  theme(legend.position = c(0.85, 0.85))+
  labs(title = "Distribution of Science Scores across Different School Types")

```
:::

::: callout-note
The boxplots of each subject indicate that the average score between private school and public school do not differ. Private schools have slightly higher scores on average. However, there is a notable insights showing that there are more outliers in public schools.

When we examine the density graph, it becomes evident that a higher degree of overlap between two types of schools. This could indicates the greater score diversity within public schools.
:::

## 5.4 Performance and Socioeconomic Status

The boxplot and ridges graph will be used to identify the relationship between socioeconomic status and the performance on each subject

::: panel-tabset
## MATHS

```{r}
ggplot(data=stu_qqq_SG_final, 
       aes(y = MATH, x=SOCIOECONOMIC_STATUS,fill=SOCIOECONOMIC_STATUS)) +
  geom_boxplot(fill = "beige")+
  scale_y_continuous(limits = c(0, 1000))+
  stat_summary(geom = "point",       
               fun.y="mean",         
               colour ="black",        
               size=2)+
  theme(legend.position = "none")

ggplot(data = stu_qqq_SG_final, 
       aes(x = MATH, y = SOCIOECONOMIC_STATUS, fill = SOCIOECONOMIC_STATUS)) +
  geom_density_ridges() +
  scale_x_continuous(limits = c(0, 1000)) +
  labs(title = "Distribution of Math Scores by Socioeconomic Status", 
       y = "Socioeconomic Status", x = "Math Score") +
  theme_ridges()+
  theme(legend.position = "none")

```

## READING

```{r}
ggplot(data=stu_qqq_SG_final, 
       aes(y = READING, x=SOCIOECONOMIC_STATUS,fill=SOCIOECONOMIC_STATUS)) +
  geom_boxplot(fill="pink")+
  scale_y_continuous(limits = c(0, 1000))+
  stat_summary(geom = "point",       
               fun.y="mean",         
               colour ="black",        
               size=2)+
  theme(legend.position = "none")

ggplot(data = stu_qqq_SG_final, 
       aes(x = READING, y = SOCIOECONOMIC_STATUS, fill = SOCIOECONOMIC_STATUS)) +
  geom_density_ridges() +
  scale_x_continuous(limits = c(0, 1000)) +
  labs(title = "Distribution of Reading Scores by Socioeconomic Status", 
       y = "Socioeconomic Status", x = "Reading Score") +
  theme_ridges()+
  theme(legend.position = "none")
```

## SCIENCE

```{r}
ggplot(data=stu_qqq_SG_final, 
       aes(y = SCIENCE, x=SOCIOECONOMIC_STATUS,fill=SOCIOECONOMIC_STATUS)) +
  geom_boxplot(fill="lightblue")+
  scale_y_continuous(limits = c(0, 1000))+
    stat_summary(geom = "point",       
               fun.y="mean",         
               colour ="black",        
               size=2)+
  theme(legend.position = "none")

ggplot(data = stu_qqq_SG_final, 
       aes(x = SCIENCE, y = SOCIOECONOMIC_STATUS, fill = SOCIOECONOMIC_STATUS)) +
  geom_density_ridges() +
  scale_x_continuous(limits = c(0, 1000)) +
  labs(title = "Distribution of Science Scores by Socioeconomic Status", 
       y = "Socioeconomic Status", x = "Science Score") +
  theme_ridges()+
  theme(legend.position = "none")
```
:::

::: callout-note
The boxplot for each subjects reveal significant difference in performance based on socioeconomic status. Students with higher socioeconomic status tend to achieve better scores. This demonstrates a positive correlation between socioeconomic status and scores. In other words, students from more advantaged backgrounds tend to perform better.

The density ridges plots show that students with more advantaged socioeconomic status cluster around higher score ranges, while those with less advantaged socioeconomic status tend to cluster around lower score ranges. This reaffirms that socioeconomic advantages are associated with better academic outcome.
:::

# 6 Conclusion

In summary,our analysis reveal some key findings. Across all three subjects, there is a similarity in score distributions. However, math consistently achieve the highest scores, while reading tends to have lowest. Additionally, female students tend to have around similar scores, while male students have greater score variability. While private schools tend to have slightly better average scores, public schools have a broader range of scores. This indicates that there is more variation in how well students in public schools perform. Lastly, the most observation is positive correlation between socioeconomic status and academic performance. Students from more advantaged backgrounds consistently achieve better scores.

Recommendation for the further investigation is to include a long term study which involve following a specific group of students from a particular point in time, such as from elementary school and collecting data on their academic performance at multiple points in their educational journey. This will provide more comprehensive insights and reveal the trend that may not appear in a short term study.
