---
title: "Hands-on Exercise 5a: Creating Ternary Plot with R"
date: February 4,2024
date-modified: "last-modified"
toc: true
execute: 
  eval: true
  echo: true
  warning: false
code-fold: true
code-summary: "Show the code"
---

# 1 Getting Started

In this exercise, we will use the following our R packages.

-   ggtern: a ggplot extension specially designed to plot ternary diagrams. The package will be used to plot static ternary plots.

-   [Plotly R](https://plotly.com/r/), R library for plotting interactive statistical graphs via plotly’s JavaScript graphing library, plotly.js. The **plotly R** libary contains the ggplotly function, which will convert **ggplot2** figures into a Plotly object.

-   [tidyverse](https://www.tidyverse.org/), a family of modern R packages specially designed to support data science, analysis and communication task.

::: panel-tabset
## Loading R package

The code chunk below uses `p_load()` of pacman package to check if these packages are installed in the computer and load them onto your working R environment.

```{r}
#| code-fold: false
pacman::p_load(plotly, ggtern, tidyverse)
```

## Importing Data

The code chunk below imports respopagsex2000to2018_tidy.csv into R environment by using `read_csv()` function of [readr](https://readr.tidyverse.org/) package.

```{r}
#| code-fold: false
pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv") 
```

## Preparing Data

Next, use the `mutate()` function of **dplyr** package to derive three new measures, namely: young, active, and old.

```{r}
#| code-fold: false
#Deriving the young, economy active and old measures
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)
```
:::

# 2 Plotting Ternary Diagram with R

# 2.1 Plotting a static ternary diagram

Use `ggtern()` function of **ggtern** package to create a simple ternary plot.

```{r}
#| code-fold: false
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()
```

```{r}
#| code-fold: false
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()+theme_rotate()
```

## 2.2 Plotting an interative ternary diagram

The code below create an interactive ternary plot using `plot_ly()` function of **Plotly R**.

```{r}
#| code-fold: false
# reusable function for creating annotation object
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "black"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

# Initiating a plotly visualization 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("red"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```
