---
title: "Hands-on Exercise 5d: Visual Multivariate Analysis with Parallel Coordinates Plot"
date: February 7,2024
date-modified: "last-modified"
toc: true
execute: 
  eval: true
  echo: true
  warning: false
code-fold: false
code-summary: "Show the code"
---

# 1 Getting Started

In this exercise, we will use the following our R packages.

-   *`ggparcoord()`* of **GGally** package to plot statistic parallel coordinates plot,

-   **parcoords** package to plot interactive parallel coordinates plots, and

-   **parallelPlot** package to plot interactive parallel coordinates plots.

::: panel-tabset
## Loading R package

The code chunk below uses `p_load()` of pacman package to check if these packages are installed in the computer and load them onto your working R environment.

```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```

## Importing Data

The code chunk below imports WHData-2018.csv into R environment by using `read_csv()` function of [readr](https://readr.tidyverse.org/) package.

```{r}
wh <- read_csv("data/WHData-2018.csv")
```
:::

# 2 Plotting Static Parallel Coordinates Plot

## 2.1 A simple parallel coordinates

Code chunk below shows a typical syntax used to plot a basic static parallel coordinates plot by using `ggparcoord()`.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12))
```

::: callout-note
Notice that only two argument namely `data` and `columns` is used. `Data` argument is used to map the data object (i.e. *wh*) and `columns` is used to select the columns for preparing the parallel coordinates plot.
:::

## 2.2 A parallel coordinates with boxplot

The basic parallel coordinates failed to reveal any meaning understanding of the World Happiness measures. In this section, `ggparcoord()` will be used to makeover the plot.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")
```

::: callout-tip
## Things to learn from the code chunk above

-   `groupColumn` argument is used to group the observations (i.e. parallel lines) by using a single variable (i.e. Region) and colour the parallel coordinates lines by region name.

-   `scale` argument is used to scale the variables in the parallel coordinate plot by using `uniminmax` method. The method univariately scale each variable so the minimum of the variable is zero and the maximum is one.

-   `alphaLines` argument is used to reduce the intensity of the line color to 0.2. The permissible value range is between 0 to 1.

-   `boxplot` argument is used to turn on the boxplot by using logical `TRUE`. The default is `FALSE`.

-   `title` argument is used to provide the parallel coordinates plot a title.
:::

## 2.3 Parallel coordinates with facet

In the code chunk below, `facet_wrap()` of **ggplot2** is used to plot 10 small multiple parallel coordinates plots. Each plot represent one geographical region such as East Asia.

```{r}
#| fig-width: 12
#| fig-height: 5
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)

```

## 2.4 Rotating x-axis text label

We can rotate axis text labels using `theme()` function in ggplot2 as shown in the code chunk below.

```{r}
#| fig-width: 15
#| fig-height: 7
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30))

```

::: callout-tip
## Things to learn from the code chunk above

-   To rotate x-axis text labels, we use `axis.text.x` as argument to `theme()` function. And we specify `element_text(angle = 30)` to rotate the x-axis text by an angle 30 degree.
:::

## 2.5 Adjusting the rotated x-axis text label

Rotating x-axis text labels to 30 degrees makes the label overlap with the plot and we can avoid this by adjusting the text location using `hjust` argument to theme’s text element with `element_text()`. We use `axis.text.x` as we want to change the look of x-axis text.

```{r}
#| fig-width: 15
#| fig-height: 7
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```

# 3 Plotting Interactive Parallel Coordinates Plot: parallelPlot methods

In this section, [**parallelPlot**](https://cran.r-project.org/web/packages/parallelPlot/) will be used to build interactive parallel coordinates plot.

## 3.1 The basic plot
The code chunk below plot an interactive parallel coordinates plot by using `parallelPlot()`.
```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))
parallelPlot(wh,
             width = 320,
             height = 250)
```

## 3.2 Rotate axis label
In the code chunk below, `rotateTitle` argument is used to avoid overlapping axis labels.

```{r}
parallelPlot(wh,
             rotateTitle = TRUE)
```

## 3.3 Changing the color scheme
We can change the default blue color scheme by using `continousCS` argument as shown in the code chunk below.
```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)
```

## 3.4 Parallel coordinates plot with histogram
In the code chunk below, `histoVisibility` argument is used to plot histogram along the axis of each variables.
```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)
```








